Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2024-11-28 paper/main.tex
Minor paper updates
Fix plots in paper
Updates
Updates paper

Understanding cause-and-effect relationships between variables is a fundamental
goal in many scientific inquiries, as it reveals the underlying mechanisms of
observed phenomena and help in determining effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to uncover these relationships
among random variables from observational data. Methods for causal discovery
have been developed in both the Directed Acyclic Graphs (DAGs) and Structural
Equation Models (SEMs) literatures, each adopting a different approach.

In the DAG literature, the focus has primarily been on developing automated
algorithms to learn causal structures from datasets. As a result of these
efforts, many algorithms for causal discovery have been proposed, such as
constraint-based methods like the PC algorithm \citep{Spirtes2001} and Fast
Causal Inference (FCI) \citep{Spirtes2000}), score-based methods like
Hill-Climb Search and Greedy Equivalence Search (GES) \citep{Chickering2002},
and continuous optimization-based methods like NO TEARS \citep{Zheng2018} and
DAGMA \citep{Bello2022}. In contrast, the SEM literature has focused more on
developing tools to assist researchers in manually constructing or modifying
models while allowing them to integrate their expert knowledge into the
model-building process. Researchers typically develop an initial model based on
their expertise and then use various tools to guide modifications to improve
the model's fit to data. This model construction process is commonly known as
\emph{Specification Search} \citep{Long1983} and uses method like modification
indices, and Wald-based tests \citep{Marcoulides2018}.

@@

\begin{enumerate}
	\item \textbf{Lack of Trust:} Although most algorithms are
		asymptotically consistent, their finite sample properties are
		not well understood. Moreover, their output can vary
		significantly depending on the chosen algorithm and their
		hyperparameters. Combined with the lack of robust methods to
		test these algorithms' performance on a given dataset, it can
		be difficult to trust their outputs. Consequently, researchers
		usually have to rely on post-hoc modifications to the model
		based on model testing and their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges before using them for analysis.
\end{enumerate}

Figure~\ref{fig:intro} shows an example of application of a causal discovery
algorithms namely PC on the Adult Income dataset. \todo[inline]{Explain}.

Due to these challenges, whether constructing the model manually or using an
algorithm, we typically have to rely on some expert knowledge to get a model
that we can use for further analysis. Hence, it is crucial to have tools to
guide this construction or modification process.

One of the ways to guide this modification is to use Conditional Independence
(CI) based model testing \cite{Ankan2023}. As DAGs imply CI statements that can
be tested in the given data using statistical CI tests, we can verify whether
the implied CIs are consistent with the data or not. If some of implied CIs are
not valid in the given dataset, we can modify the model accordingly.
Figure~\ref{fig:ci_testing} shows an example of model testing using CIs.
However, using p-values of the CI tests has a few drawbacks. Firstly, as CIs
are implied only by missing edges in the model, we can not test whether an
incorrect edge is already present in the model. For examples, in high sample
size scenarios, we might get a significant association between a pair of
variables with very small association. Secondly, the number of implied CIs can
be quite large. For example, in the moderately sized alarm network with $ 37 $
node, there are $287$ implied CIs. If we rely only on the p-values of the CI
tests, it can be difficult to figure which test to focus on.

A common method for specification search is using \emph{modification indices},
which identify potential changes to the model structure that improve the model
fit. These potential changes are then ranked by the ones that improve the fit
the most. This ranking allows the researchers to focus first on the suggestions
that can improve the model the most and while integrating their expertise to
choose the most optimal modifications.

Inspired by this ranking approach in modification indices, in this paper we
develop a method to rank the CI tests using a conditional measure of
association.  To deal with both this issues, in this paper, we develop a
measure of partial association which can be using in combination of the CI test
to rank the violations in these CI tests. On the other hand, in Structural
Equation Models (SEMs) literature, more emphasis has been placed on developing
tools to help researchers manually construct models while integrating their
domain expertise. Our proposed approach uses the measure of association to find
correlated variables in the data that are not explained by the current model.
Using the measure, we can then rank the pair of variables between which adding
an edge would improve the explanation the most. Based on this information, the
user can select a potential pair of variables and based on their expert
knowledge decide the direction of the edge.

@@

			\draw[->]  (age) to[bend right=20] (ed);
			\draw[->]  (sex) to (ed);
			\draw[->]  (age) to (hrpw);
			\draw[->]  (ed) to (hrpw);
			\draw[->]  (sex) to (hrpw);
			\draw[->]  (ed) to (income);
			\draw[->]  (hrpw) to (income);
			\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
			\draw[->]  (sex) to (income);
			\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
			\draw[->]  (sex) to (occ);	
		\end{scope}
		\begin{scope}[xshift=-3cm]
			\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
				\begin{minipage}{.2\textwidth}
					\begin{equation*}
						\begin{split}
							\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
							\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\textnormal{Age} &\ci \textnormal{Sex} \\
							\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
						\end{split}
					\end{equation*}
				\end{minipage}
				};
		\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Finish figure. Show an example of model learnt using PC algorithm $\rightarrow$ Orienting the unoriented edges $\rightarrow$  Model testing}}
		\label{fig:intro}
\end{figure}

Our main contributions in this paper are as follows:
\begin{enumerate}
	\item We present a measure of (conditional) association for mixed data
		based on canonical correlations. This measure of association
		generalizes the widely used partial correlation to mixed data
		setting.
	\item Using this partial association measure, we present a procedure to
		rank CI tests. This ranking can be used to prioritize addition
		or deletion of edges from the model.
	\item Lastly, we present both a web-browser based tool and an
		implementation in the Python package to allow researchers to
		apply this method to their own datasets.
\end{enumerate}

@@

\begin{figure}
	\centering
	\includegraphics[scale=0.4]{plots/web_tool_full.png}
	\caption{A screenshot of the web tool for constructing the model. Users
		can upload their dataset after which the tool creates an empty
		graph and shows all pair of variables which are associated in
		the model using undirected red edges with the strength of
		association represented using edge width. Users can then add
		edges to the model (shown in green). Unnecessary edges are shown in 
		black.}
\end{figure}

@@

\begin{figure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/plots/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/plots/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC and Hill-Climb Search algorithms against
		manually drawn DAGs using the assistance method. As PC and
		Hill-Climb Search return the Markov Equivalence Class (MEC), we
		use the best and worst scoring orientation of the MEC to get
		the range of values. The human drawn values are done for
	multiple accuracy values of the oracle. \todo[inline]{Replace Hill-Climb with GES and improve the plot}}
\end{figure}

@@

\begin{figure}
	\centering
	\begin{subfigure}{0.5\textwidth}
		\caption{\todo[inline]{Add the prompt used}}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[scale=0.3]{../code/plots/llm_adult.png}
		\caption{\todo[inline]{Change figure to have consistent style}}
	\end{subfigure}
\end{figure}

>>>>

Understanding cause-and-effect relationships between variables is a fundamental
goal in many scientific inquiries, as it reveals the underlying mechanisms of
observed phenomena and help in determining effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to find these relationships
among random variables from observational data. Methods for causal discovery
have been developed in both the Directed Acyclic Graphs (DAGs) and Structural
Equation Models (SEMs) literatures, each adopting a different approach.

In the DAG literature, the focus has primarily been on developing automated
algorithms to learn causal structures from datasets. As a result of these
efforts, many algorithms for causal discovery have been proposed, such as
constraint-based methods like the PC algorithm \citep{Spirtes2001} and Fast
Causal Inference (FCI) \citep{Spirtes2000}), score-based methods like
Hill-Climb Search and Greedy Equivalence Search (GES) \citep{Chickering2002},
and continuous optimization-based methods like NO TEARS \citep{Zheng2018} and
DAGMA \citep{Bello2022}. In contrast, the SEM literature has focused more
towards developing tools to assist researchers in manually constructing or
modifying models. This approach allows experts to easily integrate their expert
knowledge into the model-building process. Researchers typically develop an
initial model based on their expertise and then use various tools to guide
modifications to improve the model's fit to data. This model construction
process is commonly known as \emph{Specification Search} \citep{Long1983} and
uses method like modification indices, and Wald-based tests
\citep{Marcoulides2018}.

@@

\begin{enumerate}
	\item \textbf{Lack of Trust:} Although most algorithms are
		asymptotically consistent, their finite sample properties are
		not well understood \todo{Is there any citation for this}.
		Moreover, their output can vary significantly depending on the
		chosen algorithm and their hyperparameters. Combined with the
		lack of robust methods to test these algorithms' performance on
		a given dataset, it can be difficult to trust their outputs.
		Consequently, researchers usually have to rely on post-hoc
		modifications to the model based on model testing and their
		domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges before using them for analysis.
\end{enumerate}

Figure~\ref{fig:intro} shows an example to highlight some of these issues using 
Adult income dataset.

% Due to these challenges, whether constructing the model manually or using an
% algorithm, we typically have to rely on some expert knowledge to get a model
% that we can use for further analysis. Hence, it is crucial to have tools to
% guide this construction or modification process.

One of the common ways to guide this post-hoc modification of DAGs is to use
Conditional Independence (CI) based model testing \cite{Ankan2023}. As DAGs
imply CI statements that can be tested in the given data using statistical CI
tests, we can verify whether the implied CIs are consistent with the data or
not. If some of implied CIs are not valid in the given dataset, we can modify
the model accordingly. Figure~\ref{fig:ci_testing} shows an example of model
testing using CIs. However, this CI based testing has a few drawbacks: 1) CIs
are implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example,  in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

On the other hand, a common method for specification search is using
\emph{modification indices}, which identify potential changes to the model
structure that improve the model fit. These potential changes are then ranked
by the ones that improve the fit the most. This ranking allows researchers to
focus first on the suggestions that can improve the model the most and while
integrating their expertise to choose the most optimal modifications.

Inspired by this ranking approach in modification indices, in this paper we
develop a method to rank the CI tests. We utilize a measure of conditional
association to do this ranking. Additionally, this measure of conditional
association also allows us tackle other issues with CI based testing outlined
in the previous paragraphs.

% using a conditional measure of
% association.  To deal with both this issues, in this paper, we develop a
% measure of partial association which can be using in combination of the CI test
% to rank the violations in these CI tests. On the other hand, in Structural
% Equation Models (SEMs) literature, more emphasis has been placed on developing
% tools to help researchers manually construct models while integrating their
% domain expertise. Our proposed approach uses the measure of association to find
% correlated variables in the data that are not explained by the current model.
% Using the measure, we can then rank the pair of variables between which adding
% an edge would improve the explanation the most. Based on this information, the
% user can select a potential pair of variables and based on their expert
% knowledge decide the direction of the edge.

@@

			\draw[->]  (age) to[bend right=20] (ed);
			\draw[->]  (sex) to (ed);
			\draw[->]  (age) to (hrpw);
			\draw[->]  (ed) to (hrpw);
			\draw[->]  (sex) to (hrpw);
			\draw[->]  (ed) to (income);
			\draw[->]  (hrpw) to (income);
			\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
			\draw[->]  (sex) to (income);
			\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
			\draw[->]  (sex) to (occ);	
		\end{scope}
		\begin{scope}[xshift=-3cm]
			\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
				\begin{minipage}{.2\textwidth}
					\begin{equation*}
						\begin{split}
							\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
							\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\textnormal{Age} &\ci \textnormal{Sex} \\
							\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
						\end{split}
					\end{equation*}
				\end{minipage}
				};
		\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Generate this figure using a few different CI tests and also show the unoriented edges in some of them}}
		\label{fig:intro}
\end{figure}

Our main contributions in this paper are as follows:
\begin{enumerate}
	\item We present a measure of (conditional) association for mixed data
		based on canonical correlations. This measure of association
		generalizes the widely used partial correlation to mixed data
		setting.
	\item Using this partial association measure, we present a procedure to
		rank CI tests. This ranking can be used to prioritize addition
		or deletion of edges from the model.
	\item We present both a web-browser based tool and an implementation in
		the Python package to allow researchers to apply this method to
		their own datasets.
\end{enumerate}

@@

\begin{figure}
	\centering
	% \includegraphics[scale=0.4]{plots/web_tool_full.png}
	\caption{A screenshot of the web tool for constructing the model. Users
		can upload their dataset after which the tool creates an empty
		graph and shows all pair of variables which are associated in
		the model using undirected red edges with the strength of
		association represented using edge width. Users can then add
		edges to the model (shown in green). Unnecessary edges are shown in 
		black.}
\end{figure}

@@

\begin{figure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC and Hill-Climb Search algorithms against
		manually drawn DAGs using the assistance method. As PC and
		Hill-Climb Search return the Markov Equivalence Class (MEC), we
		use the best and worst scoring orientation of the MEC to get
		the range of values. The human drawn values are done for
	multiple accuracy values of the oracle. \todo[inline]{Replace Hill-Climb with GES and improve the plot}}
\end{figure}

\begin{figure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig4/unexplained_effect.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig4/ll.pdf}
		\caption{}
	\end{subfigure}
	\caption{Plots showing improvements in fit of the model over iterations
	on the Adult income dataset. For the expert simulation an LLM is used
	for the edge orientations. (a) Shows using total unexplained effect that is the
	sum of effect size between variables that don't have an edge between them. (b)
	Shows the log-likelihood of the overall model.
	}
\end{figure}

@@

\begin{figure}
	\centering
	\begin{subfigure}{0.5\textwidth}
		\caption{\todo[inline]{Add the prompt used}}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[scale=0.3]{../code/fig5/llm_adult.png}
		\caption{\todo[inline]{Change figure to have consistent style}}
	\end{subfigure}
\end{figure}
